# QC - bulk methylation

<!-- #region tags=[] -->
### initialize definitions
<!-- #endregion -->

```{r}
suppressMessages(suppressWarnings(source(here::here("code/init.R"))))
```

## Probe design


## Breakdown of probes

Get capture regions:

```{r}
regions <- "intervs.captPBAT_probes.ES_EB_V1"
regs_df <- gintervals.load(regions) %>% as_tibble()
```

Get capture probes:
```{r}
probes <- fread(here("data/probes_V1.csv"))
probes_span <- probes %>% select(chrom, start, end, probe_id=id) %>% gintervals.centers() %>% mutate(start = start - 150, end = end + 150)
```

Get enhancer definitions:
```{r}
enh_intervs <- define_enhancer_intervs() %cache_df% here("output/enh_intervs.csv") %>% as_tibble()
```

Get promoter definitions:

```{r}
prom_intervs <- misha.ext::get_promoters(upstream = 1e3, downstream=50) %>% as_tibble()
```

Annotate regions:
```{r}
probes_annot <- probes_span %>% 
    gintervals.neighbors1(prom_intervs) %>% 
    mutate(dist_prom = dist) %>% 
    select(-(chrom1:dist)) %>% 
    gintervals.neighbors1(enh_intervs) %>% 
    mutate(dist_enh = dist) %>% 
    select(-(chrom1:dist)) %>% 
    mutate(type = case_when(dist_prom == 0 ~ "Promoter", dist_enh == 0 ~ "Enhancer", TRUE ~ "Background"))
```

### Extended Data Figure 8B

```{r probes-breakdown-plot, fig.asp=4/5}
p <- probes_annot %>% count(type) %>% mutate(type = factor(type, levels = c("Promoter", "Enhancer", "Background"))) %>% ggplot(aes(x=type, y=n)) + geom_col() + scale_y_continuous(labels=scales::comma) + xlab("") + ylab("# of probes") + theme(aspect.ratio=1) + vertical_labs()
p
```

## WGBS E8.5

### Coverage

```{r}
wgbs_track <- "e8_5"
wgbs_df <- gextract(c(glue("{wgbs_track}.cov"), glue("{wgbs_track}.meth")), intervals="intervs.global.seq_CG", iterator = "intervs.global.seq_CG", colnames = c("cov", "meth")) %>% arrange(intervalID) %>% select(-intervalID) %>% as_tibble()
wgbs_df <- wgbs_df %>% mutate(cov = ifelse(is.na(cov), 0, cov), meth = ifelse(is.na(meth), 0, meth))
wgbs_cov_counts <- wgbs_df %>% count(cov)
med_cov <- median(wgbs_df$cov)
```

Plot:


### Extended Data Figure 8A

```{r wgbs-cov-plot, fig.asp=4/5}
p <- wgbs_cov_counts %>% 
    filter(cov > 0) %>% 
    mutate(cum = rev(cumsum(rev(n)))) %>% 
    ggplot(aes(x=cov, y=cum)) + 
        geom_line() + 
        scale_x_log10(breaks=c(1,2.5,5,10, med_cov, 50, 100, 1e3), labels = c(1,'', 5,10,med_cov,'', 100,1000)) + 
        coord_cartesian(xlim=c(1,1e3)) + 
        scale_y_continuous(labels=scales::scientific) + 
        ylab("# of CpGs") + 
        xlab("E8.5 WGBS CpG coverage") + 
        geom_vline(xintercept=med_cov, linetype="dashed", color="darkblue") + 
        theme(aspect.ratio=0.7)
p
```

### Methylation distribution on-target vs off-target


### Extended Data Figure 8C

```{r wgbs-off-vs-on-target}
wgbs_df_annot <- wgbs_df %>% 
    filter(cov >= 20) %>% 
    gintervals.neighbors1("intervs.captPBAT_probes.ES_EB_V1") %>% 
    mutate(type = ifelse(is.na(dist) | dist == 0, "Designed", "Other")) %>% 
    select(-(chrom1:dist))
p <- wgbs_df_annot %>% 
    mutate(type = factor(type, levels = c("Other", "Designed"))) %>% 
    arrange(desc(type)) %>% 
    mutate(avg = meth / cov) %>% 
    ggplot(aes(x=avg, fill=type)) + 
        geom_density(alpha=0.5) + 
        scale_fill_manual("", values=c("Designed" = "red", "Other" = "gray")) + 
        theme(aspect.ratio=1) + 
        xlab("E8.5 Avg. Meth.") + 
        ylab("Density")
p
```

## Distribution of coverage per condition

Get CpGs participating in the model generation:

The criterion is coverage >= 10 in at least 5 samples.

```{r}
cpg_meth <- calc_eb_cpg_meth(from = 0, to = 6, min_cov = 10, max_na = 5, intervals = gintervals.union("intervs.captPBAT_probes.ES_EB_V1", "intervs.captPBAT_probes.ES_EB_V2"), iterator = "intervs.global.seq_CG", cache_fn = here("output/eb_day0_to_day6_cpg_meth_all.tsv"), rm_meth_cov = FALSE)
cpg_cov <- cpg_meth %>% select(chrom, start, end, ends_with("cov"))
colnames(cpg_cov) <- gsub(".cov$", "", colnames(cpg_cov))
colnames(cpg_cov) <- gsub("_tko", "_TKO", colnames(cpg_cov))
colnames(cpg_cov) <- gsub("d0S_", "S_", colnames(cpg_cov))
```

```{r sample-defs}
samp_levels <- c("d0_3a", "d1_3a", "d2_3a", "d3_3a", "d4_3a", "d0_3b",
"d1_3b", "d2_3b", "d3_3b", "d4_3b", "d0_wt", "d1_wt",
"d2_wt", "d3_wt", "d4_wt", "d0_TKO", "d1_TKO", "d2_TKO",
"d3_TKO", "d4_TKO", "S_3a", "S_3b", "S_wt", "S_TKO")
mut_levels <- c("3a", "3b", "wt", "TKO")
```

```{r sample-defs-1}
samp_levels_all <- c("d0_3a", "d1_3a", "d2_3a", "d3_3a", "d4_3a", "d5_3a", "d6_3a", "d0_3b",
"d1_3b", "d2_3b", "d3_3b", "d4_3b", "d5_3b", "d6_3b", "d0_wt", "d1_wt",
"d2_wt", "d3_wt", "d4_wt", "d5_wt", "d6_wt", "d0_TKO", "d1_TKO", "d2_TKO",
"d3_TKO", "d4_TKO", "S_3a", "S_3b", "S_wt", "S_TKO")#, "N15_d4_3a", "N15_d5_3a", "N15_d4_3b", "N15_d5_3b", "N15_d4_wt", "N15_d5_wt")
mut_levels_all <- c("3a", "3b", "wt", "TKO")
```

```{r}
covs_df <- cpg_cov %>% 
    gather("samp_id", "cov", -(chrom:end)) %>% 
    mutate(samp_id = factor(samp_id, levels = samp_levels_all)) %>% 
    filter(!is.na(samp_id)) %>% 
    separate(samp_id, c("day", "mut"), sep="_", remove=FALSE) %>% 
    mutate(mut = factor(mut, levels = mut_levels_all)) %cache_df% here("output/cpg_cov_capt_all.tsv")
```

```{r}
covs_df %>% distinct(chrom, start, end) %>% nrow()
```

Plot coverage distribution:

### Extended Data Figure 8D

```{r coverage-distribution, fig.asp=5/12}
options(repr.plot.width = 12, repr.plot.height = 8)
p <- covs_df %>%         
    ggplot(aes(x=samp_id, y=cov, fill=mut)) + 
        geom_boxplot(outlier.shape=NA) + 
        vertical_labs() + 
        ggsci::scale_fill_npg(name='') + 
        theme(aspect.ratio=0.4) + 
        scale_y_log2(breaks = c(4, 10, 20, 50, 100, 256), limits=c(1,300)) + 
        ylab("CpG coverage\n(on target)") + 
        xlab("")
p + theme_bw() + vertical_labs() + theme(aspect.ratio=0.4) 
```

## Distribution of non-CpGs methylation per condition
<!-- #region tags=[] -->
### Extended Data Figure 8H
<!-- #endregion -->

```{r chh-distribution, fig.asp=5/12}
p <- track_stats %>%     
    filter(samp_id %in% samp_levels_all) %>% 
    mutate(mut = factor(mut, levels = mut_levels_all), 
           samp_id = factor(samp_id, levels=samp_levels_all)) %>%     
    group_by(samp_id, mut) %>% 
    summarise(CHH = mean(CHH, na.rm=TRUE)) %>% 
    ggplot(aes(x=samp_id, y=CHH, fill=mut)) + 
    geom_col() + 
    ggsci::scale_fill_npg(name='') + 
    vertical_labs() + 
    xlab('') + 
    ylab('%CHH') + 
    scale_y_continuous(labels=scales::percent) + 
    theme(aspect.ratio=0.4)
p + theme_bw() + vertical_labs() + theme(aspect.ratio=0.4) 
```

## "On target" statistics
```{r on-target-distribution}
ontar_stats <- calc_on_target_stats() %cache_df% here("data/on_target_stats.csv") %>% as_tibble()
df <- ontar_stats %>% 
    left_join(tracks_key %>% select(track = track_name, capture)) %>% 
    mutate(n_offtar = n - n_ontar) %>% 
    pivot_longer(cols=c("n_ontar", "n_offtar"), names_to="type", values_to="num") %>% 
    mutate(type = case_when(type == "n_ontar" ~ "on target", type == "n_offtar" ~ "off target")) %>%
    filter(samp_id %in% samp_levels_all) %>% 
    mutate(mut = factor(mut, levels = mut_levels_all), samp_id = factor(samp_id, levels=samp_levels_all)) %>% 
    group_by(samp_id, mut, day, type) %>% 
    summarise(num = sum(num, na.rm=TRUE) + 1)
```

<!-- #region tags=[] -->
### Extended Data Figure 8E
<!-- #endregion -->

```{r}
p_0_4 <- df %>%        
    filter(!(day %in% c("d5", "d6"))) %>%    
    ggplot(aes(x=samp_id, y=num, fill=type)) + 
        geom_col() + 
        scale_fill_manual(name="", values=c("on target" = "darkblue", "off target" = "darkgray")) + 
        vertical_labs() + 
        scale_y_continuous(labels=scales::scientific) +         
        xlab('') + 
        ylab('# of methylation calls') +         
        theme(aspect.ratio=0.4)
p_0_4 + theme_bw() + vertical_labs() + theme(aspect.ratio=0.4)
```

```{r}
p_5_6 <- df %>%        
    filter(day %in% c("d5", "d6")) %>%    
    ggplot(aes(x=samp_id, y=num, fill=type)) + 
        geom_col() + 
        scale_fill_manual(name="", values=c("on target" = "darkblue", "off target" = "darkgray")) + 
        vertical_labs() + 
        scale_y_continuous(labels=scales::scientific) +         
        xlab('') + 
        ylab('# of methylation calls')        
p_5_6 + theme_bw() + vertical_labs() + theme(aspect.ratio=0.4)
```

Note that we calculate the "on target" based only on mapped reads that had at least one CpG.





## Global methylation trend

Global methylation is calculated based on CpGs that are not: 

- promoters
- putative enhancers
- part of a capture region ("on target")


### Extended Data Figure 8G

```{r global-meth-distribution}
gmeth <- calc_global_meth() %cache_df% 
    here("output/global_meth.csv") %>%
    mutate(breaks = forcats::fct_recode(breaks, "2%-3%" = "(0.02,0.03]", "1%-2%" = "[0.01,0.02]", "8%-15%" = "(0.08,0.15]", "0%-3%" = "[0,0.03]")) %>% 
    filter(cg_num >= 1e4)
gmeth <- gmeth %>% 
    select(-cg_num) %>% 
    spread(breaks, meth) %>% 
    filter(samp_id %in% samp_levels_all) %>% 
    mutate(mut = factor(mut, levels = mut_levels_all), samp_id = factor(samp_id, levels=samp_levels_all))
p <- gmeth %>% 
    group_by(samp_id, mut) %>% 
    summarise(`0%-3%` = mean(`0%-3%`, na.rm=TRUE)) %>%     
    ggplot(aes(x=samp_id, y=`0%-3%`, fill=mut)) + geom_col() + ggsci::scale_fill_npg(name='') + vertical_labs()  + xlab('') + ylab('Avg. meth in low CpG content') + theme(aspect.ratio=0.4)
p
```

Avg. meth in high CpG content:
```{r global-meth-distribution-high-cpg-content}
p <- gmeth %>%
    group_by(samp_id, mut) %>% 
    summarise(`8%-15%` = mean(`8%-15%`, na.rm=TRUE)) %>%   
    ggplot(aes(x=samp_id, y=`8%-15%`, fill=mut)) + geom_col() + ggsci::scale_fill_npg(name='') + vertical_labs()  + xlab('') + ylab('Avg. meth in high CpG content') + theme(aspect.ratio=0.4)
p
```
